---
title: "Gender Representation in Hip-Hop Criticism"
subtitle: "An Analysis of the BBC Music Poll"
format: 
  revealjs:
    theme: dark
    transition: fade
    slide-number: true
    code-fold: true
    code-tools: true
editor: visual
---

```{r setup, include=FALSE}
library(readr)
library(dplyr)
library(ggplot2)
library(ggeffects)
library(tidyverse)

polls <- read_csv("data/polls.csv")
ranks <- read_csv("data/rankings.csv")
```

## Background

BBC Music asked **100+ critics, artists, and industry professionals** from **15 countries** to submit their five favorite hip-hop tracks.

. . .

**Scoring Algorithm**

| Place | Points |
|-------|--------|
| 1st   | 10     |
| 2nd   | 8      |
| 3rd   | 6      |
| 4th   | 4      |
| 5th   | 2      |

. . .

> Ties broken by total votes, then by first-place votes, then second, and so on.

------------------------------------------------------------------------

## Research Question

- Are female hip-hop artists **underrepresented** in critics' top five lists?
- When nominated, are they ranked **as highly** as male artists?  
- Does a critic's **country or role** predict whether they nominate female artists?
- Is underrepresentation a **bias problem**, a **pipeline problem**, or both?

------------------------------------------------------------------------

## The Data {.smaller}

::::: columns
::: {.column width="50%"}
**polls.csv** — one row per vote

| Variable          | Description       |
|-------------------|-------------------|
| `rank`            | Rank given (1–5)  |
| `title`           | Song title        |
| `artist`          | Artist name       |
| `gender`          | Gender of artist  |
| `year`            | Year released     |
| `critic_name`     | Critic's name     |
| `critic_rols`     | Critic's role     |
| `critic_country`  | Primary country   |
| `critic_country2` | Secondary country |
:::

::: {.column width="50%"}
**rankings.csv** — one row per song

| Variable  | Description      |
|-----------|------------------|
| `title`   | Song title       |
| `artist`  | Artist name      |
| `year`    | Year released    |
| `gender`  | Gender of artist |
| `points`  | Total points     |
| `n`       | Total votes      |
| `n1`–`n5` | Votes by rank    |
:::
:::::

------------------------------------------------------------------------

## Summary Statistics {.smaller}

**Average rank by gender** — lower is better (1 = best)

```{r}
polls |> 
  filter(!is.na(gender)) |>
  group_by(gender) |>
  summarise(
    `Avg Rank` = round(mean(rank), 2),
    `Total Votes` = n()
  ) |>
  knitr::kable()
```

. . .

**Rank distribution in rankings.csv**

```{r}
ranks |> 
  group_by(gender) |>
  summarise(
    `#1 Vote %` = round(sum(n1) / sum(n), 2),
    `#2 Vote %` = round(sum(n2) / sum(n), 2),
    `#3 Vote %` = round(sum(n3) / sum(n), 2),
    `#4 Vote %` = round(sum(n4) / sum(n), 2),
    `#5 Vote %` = round(sum(n5) / sum(n), 2)
  ) |>
  knitr::kable()
```

------------------------------------------------------------------------

## Key Finding: Ranking Gap

```{r}
polls |>
  filter(!is.na(gender)) |>
  group_by(gender) |>
  summarise(avg_rank = mean(rank)) |>
  ggplot(aes(x = gender, y = avg_rank, fill = gender)) +
  geom_col(width = 0.5) +
  scale_fill_viridis_d() +
  theme_minimal(base_size = 16) +
  labs(
    title = "Average Rank by Artist Gender",
    subtitle = "Lower = ranked higher (1 is best)",
    x = NULL, y = "Average Rank"
  ) +
  theme(legend.position = "none")
```

> Female artists avg **3.5** vs male artists avg **2.9**

------------------------------------------------------------------------

## Votes by Gender

```{r}
ggplot(data = polls, 
       aes(x = gender, fill = factor(rank))) + 
  geom_bar() + 
  theme_minimal(base_size = 16) +
  scale_fill_viridis_d() +
  labs(
    title = "Count of Top 5 Votes by Artist Gender",
    x = "Artist Gender",
    y = "Number of Votes",
    fill = "Rank"
  )
```

------------------------------------------------------------------------

## Trends Over Time

Are female artists gaining more nominations over time?

```{r}
year_gender <- polls |>
  group_by(year, gender) |>
  summarise(count = n(), .groups = "drop")

ggplot(year_gender, aes(x = year, y = count, color = gender)) +
  geom_point(alpha = 0.6) +
  geom_smooth(method = "lm", se = FALSE) +
  theme_minimal(base_size = 16) +
  scale_color_viridis_d() +
  labs(
    title = "Songs Nominated by Gender Over Time",
    subtitle = "1989 was the first year a mixed group appeared in the top 5",
    x = "Year", y = "Number of Songs", color = "Gender"
  )
```

------------------------------------------------------------------------

## Country Analysis

Do critics from certain countries nominate more female artists?

```{r}
country_gender <- polls |>
  filter(!is.na(critic_country), !is.na(gender)) |>
  group_by(critic_country) |>
  summarise(
    total = n(),
    ratio_female = mean(gender == "female", na.rm = TRUE)
  ) |>
  filter(total >= 5) |>
  arrange(desc(ratio_female))

ggplot(country_gender, 
       aes(x = reorder(critic_country, ratio_female), 
           y = ratio_female, fill = ratio_female)) +
  geom_col() +
  geom_hline(yintercept = mean(country_gender$ratio_female), 
             linetype = "dashed", color = "red", linewidth = 0.8) +
  coord_flip() +
  scale_fill_viridis_c() +
  theme_minimal(base_size = 14) +
  labs(
    title = "Female Artist Nominations by Critic's Country",
    subtitle = "Red dashed line = overall average | Countries with <5 votes excluded",
    x = NULL, y = "Proportion of female artist nominations",
    fill = "Proportion"
  )
```

------------------------------------------------------------------------

## Critic Role Analysis

Does a critic's professional role predict whether they nominate female artists?

```{r}
role_female <- polls |>
  filter(!is.na(critic_rols), !is.na(gender)) |>
  group_by(critic_rols) |>
  summarise(
    total = n(),
    ratio_female = mean(gender == "female", na.rm = TRUE)
  ) |>
  filter(total >= 5)

ggplot(role_female, 
       aes(x = reorder(critic_rols, ratio_female), 
           y = ratio_female, fill = ratio_female)) +
  geom_col() +
  geom_hline(yintercept = mean(role_female$ratio_female),
             linetype = "dashed", color = "red", linewidth = 0.8) +
  coord_flip() +
  scale_fill_viridis_c() +
  theme_minimal(base_size = 14) +
  labs(
    title = "Female Artist Nominations by Critic Role",
    subtitle = "Red dashed line = overall average",
    x = NULL, y = "Proportion of female artist nominations",
    fill = "Proportion"
  )
```

------------------------------------------------------------------------

## Predictive Model

**Does artist gender predict how highly a song is ranked?**

```{r}
polls <- polls |> mutate(gender = factor(gender))
model1 <- lm(rank ~ gender, data = polls)
summary(model1)
```

------------------------------------------------------------------------

## Model Visualization

```{r}
pred <- ggpredict(model1, terms = c("gender")) |>
  as.data.frame() |>
  as_tibble()

ggplot(data = polls, aes(x = gender, y = rank)) +
  geom_point(aes(fill = gender),
             position = position_jitter(width = 0.25, height = 0.12),
             alpha = 0.3, shape = 21) +
  geom_point(data = pred,
             aes(x = x, y = predicted, fill = x),
             size = 5, shape = 21) +
  theme_minimal(base_size = 16) +
  scale_fill_viridis_d() +
  labs(
    title = "Does Artist Gender Affect Song Ranking?",
    subtitle = "Large points = model predicted means",
    x = "Artist Gender", y = "Rank (1 = Best)",
    fill = "Gender"
  )
```

------------------------------------------------------------------------

## Conclusions

- **Underrepresentation is real:** ~70–80% of nominated artists are male, matching broader industry ratios
- **Ranking gap exists:** Female artists average rank 3.5 vs. 2.9 for male artists
- **Mixed groups fall in between** (avg 3.15) — group membership partially buffers the penalty
- **Country and role matter:** Where a critic is from influences whether they nominate women
- **Root cause question:** Is this bias, a pipeline problem, or both?

