---
title: "Final Presentation"
format: revealjs
editor: visual
---

## Data Description

BBC Music polled more than 100 critics, artists, and music industry professionals from 15 countries, asking each participant to submit their five favorite hip-hop tracks. The results were compiled into a definitive ranked list using a straightforward points-based algorithm. Songs received 10 points for a first-place vote, eight for second, and so on down to two points for fifth place, with the highest total winning.

------------------------------------------------------------------------

## polls.cvs
| item            | description                |
|-----------------|----------------------------|
| rank            | Rank given by voter (1-5)  |
| title           | Title of song              |
| artist          | Artist                     |
| gender          | Gender of artist           |

## polls.cvs
| item            | description                |
|-----------------|----------------------------|
| year            | Year song released         |
| critic_name     | Name of critic             |
| critic_rols     | Critic's role              |
| critic_country  | Critic's primary country   |
| critic_country2 | Critic's secondary country |

## ranks.cvs

-   ID: ID of song\
-   title: Title of song\
-   artist: Artist's name\
-   year: Year song released\
-   gender: Gender of artist/group\
-   points: Total points awarded\
-   n: Total votes\
-   n1–n5: Votes by rank


```{r message = FALSE}
library(readr)
polls <- read_csv("data/polls.csv")
ranks <- read_csv("data/rankings.csv")
```

## Summary Statistics
```{r}
library(dplyr)
polls |> group_by(gender) |>
  summarise(mean = mean(rank))
```


```{r}
library(dplyr)
ranks |> group_by(gender) |>
  summarise(ratio_n1 = sum(n1) / sum(n),
            ratio_n2 = sum(n2) / sum(n),
            ratio_n3 = sum(n3) / sum(n),
            ratio_n4 = sum(n4) / sum(n),
            ratio_n5 = sum(n5) / sum(n))
```

# Critics Analysis

```{r}
## see the ratio of women in each critics top five
library(dplyr)
polls2 <- polls |> group_by(critic_name) |>
  summarise(sum = sum(gender == "female", na.rm = TRUE),
            ratio_women_top5 = mean(gender == "female", na.rm = TRUE),
            ratio_men_top5 = mean(gender == "male", na.rm=TRUE))
```

# Visualization

## Votes by Gender

**prompt (claude):** code + debug, added factor to fill to aes:

```{r}
library(ggplot2)
ggplot(data = polls, 
       aes(x = gender, 
           fill = factor(rank))) + 
  geom_bar() + 
  theme_minimal() +
  scale_fill_viridis_d() +
  labs(title = "Count of Top 5 Vote by Gender")
```
