---
title: "Final Project"
format: html
---
Description: (directly from TidyTuesday)
" Earlier this year, BBC Music asked more than 100 critics, artists, and other music industry folks from 15 countries for their five favorite hip-hop tracks. Then they broke down the results of the poll into one definitive list. But BBC Music didn't just publish a best-of list, they also published the complete poll results and a description of the simple algorithm they ranked the songs with. - Simon Jockers

We awarded 10 points for first ranked track, eight points for second ranked track, and so on down to two points for fifth place. The song with the most points won. We split ties by the total number of votes: songs with more votes ranked higher. Any ties remaining after this were split by first place votes, followed by second place votes and so on: songs with more critics placing them at higher up the lists up ranked higher. -- BBC Music "

**polls**
variable	class	description
rank	double	Rank given by voter (1-5)
title	character	Title of song
artist	character	Artist
gender	character	Gender of artist
year	double	Year song released
critic_name	character	Name of critic
critic_rols	character	Critic's role
critic_country	character	Critic's primary country
critic_country2	character	Critic's secondary country

**ranks**
variable	class	description
ID	double	ID of song
title	character	Title of song
artist	character	Artist's name
year	double	Year song released
gender	character	Gender of artist/group
points	double	Total points awarded
n	double	Total votes (regardless of position)
n1	double	Number of votes as #1
n2	double	Number of votes as #2
n3	double	Number of votes as #3
n4	double	Number of votes as #4
n5	double	Number of votes as #5

```{r message = FALSE} 
library(readr)
polls <- read_csv("data/polls.csv")
ranks <- read_csv("data/rankings.csv")
```

```{r}
head(polls)
```

```{r}
head(ranks)
```


```{r}
library(dplyr)
polls |> group_by(gender) |>
  summarise(mean = mean(rank))
```

```{r}
library(dplyr)
ranks |> group_by(gender) |>
  summarise(ratio_n1 = sum(n1) / sum(n),
            ratio_n2 = sum(n2) / sum(n),
            ratio_n3 = sum(n3) / sum(n),
            ratio_n4 = sum(n4) / sum(n),
            ratio_n5 = sum(n5) / sum(n))
```

```{r}
## see the ratio of women in each critics top five
library(dplyr)
polls2 <- polls |> group_by(critic_name) |>
  summarise(sum = sum(gender == "female", na.rm = TRUE),
            ratio_women_top5 = mean(gender == "female", na.rm = TRUE),
            ratio_men_top5 = mean(gender == "male", na.rm=TRUE))
```


**prompt (claude):** code + debug, added factor to fill to aes:
```{r}
library(ggplot2)
ggplot(data = polls, 
       aes(x = gender, 
           fill = factor(rank))) + 
  geom_bar() + 
  theme_minimal() +
  scale_fill_viridis_d() +
  labs(title = "Count of Top 5 Vote by Gender")
```


```{r}
library(dplyr)
library(ggplot2)

country_gender <- polls |>
  filter(!is.na(critic_country), !is.na(gender)) |>
  group_by(critic_country) |>
  summarise(
    total = n(),
    ratio_female = mean(gender == "female", na.rm = TRUE)
  ) |> 
  arrange(desc(ratio_female))


```


**ai**: asked for plot reccommendations with this code: 
ggplot(country_gender, aes(x = reorder(critic_country, ratio_female), y = ratio_female, fill = ratio_female)) +
  geom_col() +
  scale_fill_viridis_c() +
  theme_minimal() +
  labs(
    title = "Proportion of Female Artist Nominations by Critic's Country",
    subtitle = "Dashed line = overall average | Countries with < 5 votes excluded",
    x = NULL,
    y = "Proportion of votes for female artists"
  )

took this: +
  geom_hline(yintercept = mean(country_gender$ratio_female), 
             linetype = "dashed", color = "blue", linewidth = 0.8) +
             
```{r}
ggplot(country_gender, aes(x = reorder(critic_country, ratio_female), y = ratio_female, fill = ratio_female)) +
  geom_col() +
  geom_hline(yintercept = mean(country_gender$ratio_female), 
             linetype = "dashed", color = "red", linewidth = 0.8) +
  coord_flip() +
  scale_fill_viridis_c() +  # continuous scale since ratio_female is numeric
  theme_minimal() +
  labs(
    title = "Proportion of Female Artist Nominations by Critic's Country",
    subtitle = "Dashed line = overall average | Countries with < 5 votes excluded",
    x = NULL,
    y = "Proportion of votes for female artists"
  )
```























