---
title: "Final Project"
format: pdf
---

# Introduction

Description: (directly from Tidy Tuesday)

"Earlier this year, BBC Music asked more than 100 critics, artists, and other music industry folks from 15 countries for their five favorite hip-hop tracks. Then they broke down the results of the poll into one definitive list. But BBC Music didn't just publish a best-of list, they also published the complete poll results and a description of the simple algorithm they ranked the songs with. - Simon Jockers

We awarded 10 points for first ranked track, eight points for second ranked track, and so on down to two points for fifth place. The song with the most points won. We split ties by the total number of votes: songs with more votes ranked higher. Any ties remaining after this were split by first place votes, followed by second place votes and so on: songs with more critics placing them at higher up the lists up ranked higher. -- BBC Music "

---

# Data Description

## polls

- rank: Rank given by voter (1-5)  
- title: Title of song  
- artist: Artist  
- gender: Gender of artist  
- year: Year song released  
- critic_name: Name of critic  
- critic_rols: Critic's role  
- critic_country: Critic's primary country  
- critic_country2: Critic's secondary country  

## ranks

- ID: ID of song  
- title: Title of song  
- artist: Artist's name  
- year: Year song released  
- gender: Gender of artist/group  
- points: Total points awarded  
- n: Total votes  
- n1–n5: Votes by rank  

---

# Load Data

```{r message = FALSE} 
library(readr)
polls <- read_csv("data/polls.csv")
ranks <- read_csv("data/rankings.csv")
```

# Summary Statistics

```{r}
library(dplyr)
polls |> group_by(gender) |>
  summarise(mean = mean(rank))
```

```{r}
library(dplyr)
ranks |> group_by(gender) |>
  summarise(ratio_n1 = sum(n1) / sum(n),
            ratio_n2 = sum(n2) / sum(n),
            ratio_n3 = sum(n3) / sum(n),
            ratio_n4 = sum(n4) / sum(n),
            ratio_n5 = sum(n5) / sum(n))
```

# Critics Analysis

```{r}
## see the ratio of women in each critics top five
library(dplyr)
polls2 <- polls |> group_by(critic_name) |>
  summarise(sum = sum(gender == "female", na.rm = TRUE),
            ratio_women_top5 = mean(gender == "female", na.rm = TRUE),
            ratio_men_top5 = mean(gender == "male", na.rm=TRUE))
```

# Visualization

## Votes by Gender

**prompt (claude):** code + debug, added factor to fill to aes:
```{r}
library(ggplot2)
ggplot(data = polls, 
       aes(x = gender, 
           fill = factor(rank))) + 
  geom_bar() + 
  theme_minimal() +
  scale_fill_viridis_d() +
  labs(title = "Count of Top 5 Vote by Gender")
```

# Country Analysis

```{r}
library(dplyr)
library(ggplot2)

country_gender <- polls |>
  filter(!is.na(critic_country), !is.na(gender)) |>
  group_by(critic_country) |>
  summarise(
    total = n(),
    ratio_female = mean(gender == "female", na.rm = TRUE)
  ) |> 
  arrange(desc(ratio_female))
```

**ai**: asked for plot reccommendations with this code: 
ggplot(country_gender, aes(x = reorder(critic_country, ratio_female), y = ratio_female, fill = ratio_female)) +
  geom_col() +
  scale_fill_viridis_c() +
  theme_minimal() +
  labs(
    title = "Proportion of Female Artist Nominations by Critic's Country",
    subtitle = "Dashed line = overall average | Countries with < 5 votes excluded",
    x = NULL,
    y = "Proportion of votes for female artists"
  )

took this: +
  geom_hline(yintercept = mean(country_gender$ratio_female), 
             linetype = "dashed", color = "blue", linewidth = 0.8) +
             
added: +
  geom_hline(yintercept = mean(country_gender$ratio_female), 
             linetype = "dashed", color = "red", linewidth = 0.8) +
             

```{r}
ggplot(country_gender, aes(x = reorder(critic_country, ratio_female), y = ratio_female, fill = ratio_female)) +
  geom_col() +
  geom_hline(yintercept = mean(country_gender$ratio_female), 
             linetype = "dashed", color = "red", linewidth = 0.8) +
  coord_flip() +
  scale_fill_viridis_c() +
  labs(
    title = "Proportion of Female Artist Nominations by Critic's Country",
    subtitle = "Dashed line = overall average | Countries with < 5 votes excluded",
    x = NULL,
    y = "Proportion of votes for female artists"
  ) + theme_minimal()
```

## 4/20
*google search* is new zealand more progressive with women's rights?
*result* Yes, New Zealand is considered highly progressive with women's rights, famously becoming the first country to grant women the right to vote in 1893. It boasts high levels of female political representation—nearly 50% in parliament—and strong legal frameworks for gender equality. However, challenges remain, including a gender pay gap and high rates of domestic violence. 

*google search* is new zealand more progressive with women's rights in 1990 to 2000
*result* Yes, New Zealand was considered significantly more progressive with women's rights in the 1990–2000 period, characterized by a rapid shift towards top-tier political leadership and the strengthening of legislative protection for women. This decade solidified the country's earlier reputation for equality by breaking the highest "glass ceiling" in politics and reforming domestic laws.


# Predictive Model

-- predictive model next:
Does artist gender affect how highly songs are ranked?
**not going to use most likely**

```{r}
polls <- polls |>
  mutate(gender = factor(gender))
```

```{r}
model1 <- lm(rank ~ gender, data = polls)
summary(model1)
```
- sourced from section 6 of notes
```{r}
library(tidyverse)
library(ggeffects)

pred <- ggpredict(model1, terms = c("gender")) |>
  as.data.frame() |>
  as_tibble()
```

# Model Visualization

```{r}
ggplot(data = polls, aes(x = gender, y = rank)) +
  geom_point(aes(fill = gender),
             position = position_jitter(width = 0.25, height = 0.12),
             alpha = 0.3,
             shape = 21) +
  geom_point(data = pred,
             aes(x = x, y = predicted, fill = x),
             size = 5,
             shape = 21) +
  theme_minimal() +
  scale_fill_viridis_d() + labs(
    title = "Does Artist Gender Affect Song Ranking?",
    x = "Artist Gender",
    y = "Rank (1 = Best)"
  )
```

# Trends Over Time

-- did peer review with gavin
-- does number of female artists increase with time, does number of male artists increase over time, does number of mixed artists increase over time. 

-- look at time vs count, trend line over time

```{r}
library(dplyr)

year_gender <- polls |>
  group_by(year, gender) |>
  summarise(count = n(), .groups = "drop")
```

```{r}
library(ggplot2)

ggplot(year_gender, aes(x = year, y = count, fill = gender)) +
  geom_point(aes(color = gender)) +
  geom_smooth(aes(color = gender), method = "lm", se = FALSE) +
  theme_minimal() +
  scale_fill_viridis_d() +
  scale_color_viridis_d() +
  labs(
    title = "Trends in Number of Songs by Artist Gender Over Time",
    x = "Year",
    y = "Number of Songs"
  )
```

# Reflection

Are there really more male songs, or just more male songs in these ratings?

## Claude: "from 1979 to 2019, how many male vs female music artists were there"
*Roughly 70–80% male vs. 20–30% female artists in popular music over time.*

## Why is there a random spike in female versus male artists?

## Claude: "what was going on in music from 1992 to 1998"
From 1992 to 1998, music was defined by the rise of hip-hop into the mainstream, the dominance of alternative rock, and highly commercial pop and R&B, making it one of the most influential and male-dominated eras in modern music history.

## 4/20/26
*ai* 
*prompt* here is what i have (my code), what else can i explore
*used* Does a critic's role (journalist, artist, producer, etc.) predict whether they nominate female artists? You have critic_rols unused so far.

```{r}
library(dplyr)
library(ggplot2)

role_female <- polls |>
  filter(!is.na(critic_rols), !is.na(gender)) |>
  group_by(critic_rols) |>
  summarise(
    total = n(),
    ratio_female = mean(gender == "female", na.rm = TRUE)
  ) |>
  filter(total >= 5)  # drop rare roles

ggplot(role_female, aes(x = reorder(critic_rols, ratio_female), 
                         y = ratio_female, fill = ratio_female)) +
  geom_col() +
  geom_hline(yintercept = mean(role_female$ratio_female),
             linetype = "dashed", color = "red", linewidth = 0.8) +
  coord_flip() +
  scale_fill_viridis_c() +
  theme_minimal() +
  labs(title = "Female Artist Nominations by Critic Role",
       subtitle = "Dashed line = overall average",
       x = NULL, y = "Proportion of female artist nominations")
```

## shiny app
this is my favorite visualization so far:

```{r}
ggplot(data = polls, aes(x = gender, y = rank)) +
  geom_point(aes(fill = gender),
             position = position_jitter(width = 0.25, height = 0.12),
             alpha = 0.3,
             shape = 21) +
  geom_point(data = pred,
             aes(x = x, y = predicted, fill = x),
             size = 5,
             shape = 21) +
  theme_minimal() +
  scale_fill_viridis_d() + labs(
    title = "Does Artist Gender Affect Song Ranking?",
    x = "Artist Gender",
    y = "Rank (1 = Best)"
  )
```


```{r}
library(shiny)
# inputs be year, country, 
# produce graph and table
# locked facet by gender
ui <- fluidPage(
  sidebarLayout(
    sidebarPanel(
      sliderInput("year", "Year Range",
                  min = 1979,
                  max = 2019,
                  value = c(1979, 2019),
                  sep = ""),
      radioButtons("variable_choose",
                   label = "Choose a Variable: ",
                   choices = c("critic_country", "gender"))
    ),                        # close sidebarPanel
    mainPanel(
      plotOutput("plot")
    )                         # close mainPanel
  )                           # close sidebarLayout
)                             # close fluidPage
server <- function(input, output, session) {
  
  filtered <- reactive({
    polls |>
      filter(year >= input$year[1], year <= input$year[2])
  })
  
output$plot <- renderPlot({
  ggplot(filtered(), aes(x = .data[[input$variable_choose]], y = rank, color = gender)) +
    geom_jitter(alpha = 0.4, width = 0.25, height = 0.1) +
    scale_color_viridis_d() +
    theme_minimal() +
    labs(title = paste("Votes by", input$variable_choose),
         x = input$variable_choose,
         y = "Rank") +
    theme(axis.text.x = element_text(angle = 45, hjust = 1))
})
}
shinyApp(ui, server)
```


How much of the distribution of top five ranks can we blame gender bias on versus blame the ration of male to female artists? Or is the root of the problem the lack of encouragement in female artists (underlying problems). did female artists increase over time?


```{r}
## When female artists were nominated, were they ranked fairly?
polls |>
  filter(!is.na(gender)) |>
  group_by(gender) |>
  summarise(avg_rank = mean(rank))
```
Underrepresentation: industry is ~70-80% male, polls likely match that
Ranking bias: female artists are nominated less and ranked lower when they are — avg rank 3.5 vs 2.9
Mixed sitting in the middle (3.15), groups with women aren't penalized as much as solo female artists






